Converting LaTeX R Journal Articles into Rmarkdown using texor and rebib

### 2023-07-26 texor.Rmd
documented up to pre-processing

The decision to convert to R Markdown format means that the output to pdf and HTML will depend on Pandoc [@pandoc]. 
Pandoc is a versatile document conversion program written in Haskell that is core to numerous documentation systems, including Rmarkdown and Quarto. Pandoc first converts a document into an abstract syntax tree. From this, it can convert to a different format, including custom ones. Although creating bespoke writers/readers in `Lua` can be challenging, pandoc includes filters that allow users to customize the output generated based on matching patterns plus some logic to modify them.

**DC: Can you explain what a Lua filter is?**

The \CRANpkg{texor} package uses these filters heavily, even for simple tasks like choosing which supporting figures to copy. A `Lua` filter constructs a list of image paths, stores them momentarily and the package uses this data to copy images. However, \CRANpkg{texor} also includes many pre-processing functions that employ regular expressions. This is needed to convert some text and LaTeX commands that are not recognized or handled by Pandoc.

**DC: Add a diagram here showing flow of work, just pre-process .tex -> pandoc -> lua ???**

**Maybe more details on the supporting packages here: LaTeX, Regex, pandoc building blocks of the packages, methods used in them [MOW]**

@@

LaTeX is very descriptive and allows unfettered levels of customization to authors, which poses an issue during conversions. RMarkdown is quite restrictive by  syntax though you can always embed HTML to expand the feature set. Although 
pandoc does most of the heavy lifting, there are some instances where even  filters cannot assist. A good demonstration of the issue in hand is code environments.

A limitation of pandoc is that it only works with a "verbatim" environment. However, in "RJournal",  we have multiple custom code environments which are not supported or defined due to lower-level tex limitations in pandoc. So the code  environments are eliminated during the conversion process. Including a Lua filter to find these elements and redefine them as verbatim environments would also be  ineffective. If the code environment is not verbatim, then pandoc will also try to process the actual code content as LaTeX commands and will lose out on details in some cases.

**Be sure to mention the hard parts**

## Post-Processing : Lua Filters

>>>>

The decision to convert to R Markdown format means that the final output to pdf and HTML will depend on Pandoc [@pandoc].  Pandoc is a versatile document conversion program written in Haskell that is core to numerous documentation systems, including R Markdown and Quarto. Pandoc first converts a document into an abstract syntax tree. From this, it can convert to a different format, including custom ones. <!-- Although creating bespoke writers/readers in `Lua` can be challenging, pandoc includes filters that allow users to customize the output generated based on matching patterns plus some logic to modify them. -->

Pandoc can be used to do the conversion from LaTeX to R Markdown also. However, additional pre-processing needs to be done to handle special R Journal LaTeX styling. And further post-processing needs to be done to handle specific R Journal R Markdown styling. The \CRANpkg{texor} package contains functionality to handle this pre- and postprocessing of the document.

<!-- uses these filters heavily, even for simple tasks like choosing which supporting figures to copy. A `Lua` filter constructs a list of image paths, stores them momentarily and the package uses this data to copy images. However, \CRANpkg{texor} also includes many pre-processing functions that employ regular expressions. This is needed to convert some text and LaTeX commands that are not recognized or handled by Pandoc. -->

**DC: Add a diagram here showing flow of work, just pre-process .tex -> pandoc -> lua ; rebib operations???**

<!-- **Maybe more details on the supporting packages here: LaTeX, Regex, pandoc building blocks of the packages, methods used in them [MOW]**-->

@@

LaTeX is very descriptive language, that allows authors substantial freedom for customization. Markdown [@markdown], on which R Markdown is based, is more restrictive and was born to make it easier to create web pages without the distraction of a gazillion HTML tags. The purpose of Markdown is that it allows the author to focus on writing, without format cluttering the text.

While pandoc can do most of the heavy-lifting, it cannot cope with all the freedom with which LaTeX documents are written. An example of this is with formatting of code. Pandoc only handles the `verbatim` environment, but there are many ways to format code in LaTeX, and the R Journal template has a special `\code{}` command. If the code environment is not verbatim, then pandoc will also try to process the actual code content as LaTeX commands and will likely lose details. It is better to convert these synonyms into `verbatim` environments this prior to passing the document to pandoc.

The functions in \CRANpkg{texor} that handle the pre-processing using regular expressions are:

- `stream_editor()`: operates like the `sed` function in unix [@unix] and allows generic text pattern matching and replacing.
- `patch_code_env()`: replaces the common code environments, `code`, `example`, `Sin`, `Sout`, `Scode`, `Sinput`, `smallverbatim`, `boxedverbatim`, `smallexample` with `verbatim`.
- `patch_equations()`: coordinates various equation environments.
- `patch_figure_env()`: coordinates various figure environments.
- `patch_table_env()`: coordinates table environments.

These functions are verbose and describe all the changes being made. They also create a backup of the original file before making the changes.

<!-- by syntax though you can always embed HTML to expand the feature set. Although pandoc does most of the heavy lifting, there are some instances where even  filters cannot assist. A good demonstration of the issue in hand is code environments.-->

<!-- A limitation of pandoc is that it only works with a "verbatim" environment. However, in "RJournal",  we have multiple custom code environments which are not supported or defined due to lower-level tex limitations in pandoc. So the code  environments are eliminated during the conversion process. Including a Lua filter to find these elements and redefine them as verbatim environments would also be  ineffective. If the code environment is not verbatim, then pandoc will also try to process the actual code content as LaTeX commands and will lose out on details in some cases.-->

<!--**Be sure to mention the hard parts**-->

## Post-processing using Lua filters

### 2023-07-26 texor.tex
documented up to pre-processing

The decision to convert to R Markdown format means that the output to pdf and HTML will depend on Pandoc (MacFarlane 2023).
Pandoc is a versatile document conversion program written in Haskell that is core to numerous documentation systems, including Rmarkdown and Quarto. Pandoc first converts a document into an abstract syntax tree. From this, it can convert to a different format, including custom ones. Although creating bespoke writers/readers in \texttt{Lua} can be challenging, pandoc includes filters that allow users to customize the output generated based on matching patterns plus some logic to modify them.

\textbf{DC: Can you explain what a Lua filter is?}

The \CRANpkg{texor} package uses these filters heavily, even for simple tasks like choosing which supporting figures to copy. A \texttt{Lua} filter constructs a list of image paths, stores them momentarily and the package uses this data to copy images. However, \CRANpkg{texor} also includes many pre-processing functions that employ regular expressions. This is needed to convert some text and LaTeX commands that are not recognized or handled by Pandoc.

\textbf{DC: Add a diagram here showing flow of work, just pre-process .tex -\textgreater{} pandoc -\textgreater{} lua ???}

\textbf{Maybe more details on the supporting packages here: LaTeX, Regex, pandoc building blocks of the packages, methods used in them {[}MOW{]}}

@@

LaTeX is very descriptive and allows unfettered levels of customization to authors, which poses an issue during conversions. RMarkdown is quite restrictive by syntax though you can always embed HTML to expand the feature set. Although
pandoc does most of the heavy lifting, there are some instances where even filters cannot assist. A good demonstration of the issue in hand is code environments.

A limitation of pandoc is that it only works with a ``verbatim'' environment. However, in ``RJournal'', we have multiple custom code environments which are not supported or defined due to lower-level tex limitations in pandoc. So the code environments are eliminated during the conversion process. Including a Lua filter to find these elements and redefine them as verbatim environments would also be ineffective. If the code environment is not verbatim, then pandoc will also try to process the actual code content as LaTeX commands and will lose out on details in some cases.

\textbf{Be sure to mention the hard parts}

\hypertarget{post-processing-lua-filters}{%
\subsection{Post-Processing : Lua Filters}\label{post-processing-lua-filters}}

@@

@@

>>>>

The decision to convert to R Markdown format means that the final output to pdf and HTML will depend on Pandoc (MacFarlane 2023). Pandoc is a versatile document conversion program written in Haskell that is core to numerous documentation systems, including R Markdown and Quarto. Pandoc first converts a document into an abstract syntax tree. From this, it can convert to a different format, including custom ones.

Pandoc can be used to do the conversion from LaTeX to R Markdown also. However, additional pre-processing needs to be done to handle special R Journal LaTeX styling. And further post-processing needs to be done to handle specific R Journal R Markdown styling. The \CRANpkg{texor} package contains functionality to handle this pre- and postprocessing of the document.

\textbf{DC: Add a diagram here showing flow of work, just pre-process .tex -\textgreater{} pandoc -\textgreater{} lua ; rebib operations???}

@@

LaTeX is very descriptive language, that allows authors substantial freedom for customization. Markdown (Gruber 2002), on which R Markdown is based, is more restrictive and was born to make it easier to create web pages without the distraction of a gazillion HTML tags. The purpose of Markdown is that it allows the author to focus on writing, without format cluttering the text.

While pandoc can do most of the heavy-lifting, it cannot cope with all the freedom with which LaTeX documents are written. An example of this is with formatting of code. Pandoc only handles the \texttt{verbatim} environment, but there are many ways to format code in LaTeX, and the R Journal template has a special \texttt{\textbackslash{}code\{\}} command. If the code environment is not verbatim, then pandoc will also try to process the actual code content as LaTeX commands and will likely lose details. It is better to convert these synonyms into \texttt{verbatim} environments this prior to passing the document to pandoc.

The functions in \CRANpkg{texor} that handle the pre-processing using regular expressions are:

\begin{itemize}
\tightlist
\item
  \texttt{stream\_editor()}: operates like the \texttt{sed} function in unix (Ritchie and Thompson 1978) and allows generic text pattern matching and replacing.
\item
  \texttt{patch\_code\_env()}: replaces the common code environments, \texttt{code}, \texttt{example}, \texttt{Sin}, \texttt{Sout}, \texttt{Scode}, \texttt{Sinput}, \texttt{smallverbatim}, \texttt{boxedverbatim}, \texttt{smallexample} with \texttt{verbatim}.
\item
  \texttt{patch\_equations()}: coordinates various equation environments.
\item
  \texttt{patch\_figure\_env()}: coordinates various figure environments.
\item
  \texttt{patch\_table\_env()}: coordinates table environments.
\end{itemize}

These functions are verbose and describe all the changes being made. They also create a backup of the original file before making the changes.

\hypertarget{post-processing-using-lua-filters}{%
\subsection{Post-processing using Lua filters}\label{post-processing-using-lua-filters}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-markdown}{}}%
Gruber, John. 2002. {``Markdown.''} https://daringfireball.net/projects/markdown/.

@@

\leavevmode\vadjust pre{\hypertarget{ref-unix}{}}%
Ritchie, O. M., and K. Thompson. 1978. {``The UNIX Time-Sharing System.''} \emph{The Bell System Technical Journal} 57 (6): 1905--29. \url{https://doi.org/10.1002/j.1538-7305.1978.tb02136.x}.
